A geometric relationship of $F_2, F_3$ and $F_4$-statistics with Principal Component Analysis

### 2022-02-08 fstats_pca_lewontin.tex
local revs

\subsubsection{Missing data}
The matrix decompositions underlying PCA assume complete data, and thus cannot be used when some data is missing \citep{jollife2013}.  As missing data is a very common practical problem, there is a large number of algorithms for imputing missing data. The simplest approach is to replace missing data with zeros \citep[as implemented e.g. in][]{patterson2006}, but more sophisticated algorithms exist to ``learn'' the missing values from surrounding data \citep[e.g.][]{hastie2015, meisner2018, meisner2021}. In contrast, missing data in $F$-statistics is most commonly handled by estimating a standard error by resampling along the genome \citep{patterson2012}, and so missing data results in larger standard error.

@@

\subsubsection{Individual vs. population-based analyses}
The final issue is that PCA is commonly run on individual-based data, whereas $F$-statistics often group individuals into populations.However, population-based PCA has been the default in the past \citep{cavalli-sforza1994}, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}. Often, an individual-based PCA is used to justify grouping individuals into populations; i.e. individuals that form a tight cluster on a PCA-plot have similar relationships to everyone else in the data set, and can thus be treated as a unit of analysis. Thus, if the assumptions are satisfied, $F$-statistics for individual-based and population-based analyses are expected to be very similar. PCA, on the other hand, is strongly impacted by the number of individuals from each population \citep[e.g.][]{mcvean2009}; as each individual is weighted equally, variation related to populations with many samples will be overrepresented on the first PCs.

\subsubsection{Summary}
The version of PCA used here differs from that proposed by \cite{patterson2006}, and thus some care will be required to directly extend the interpretations developed here to individual-based PCAs. However, the differences are largely due to conventions, and particularly for studies where the description of population structure is a major focus, results might be easier to interpret if conventions regarding missing data, normalization and estimation of allele frequencies are used consistently between $F$-statistics and PCA.

\subsection{The Apportionment of Human Diversity}
Most genetic variation in humans is shared between all of us, but the around 15\% that can be explained by population structure can be leveraged to study our history and diversity in great detail \citep{dobzhansky1972, cavalli-sforza1964, reich2018a}. For some data sets it is possible to predict an individuals' origin at a resolution of a few hundred kilometers \citep{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.

@@

From a practical perspective, formulating hypotheses and designing studies in terms of discrete populations with ``uniform'' genetic backgrounds is often sensible, as it enables  e.g. prediction of phenotypes \citep{berg2019, yair2021}, inference of demographic parameters, and schematic models of human genetic history \citep{patterson2012}. This is also the case for $F$-statistics, which are motivated by trees. They assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
However, these simplifications do come at a cost, both in terms of model violations that may invalidate statistical results, and in terms of deemphasizing people that do not rigidly fall into predefined genetic groups.

>>>>

\subsubsection{Missing data}
The matrix decompositions underlying PCA assume complete data, and thus cannot be used when some data is missing \citep{jolliffe2013}.  As missing data is a very common practical problem, there is a large number of algorithms for imputing missing data. The simplest approach is to replace missing data with zeros \citep[as implemented e.g. in][]{patterson2006}, but more sophisticated algorithms exist to ``learn'' the missing values from surrounding data \citep[e.g.][]{hastie2015, meisner2018, meisner2021}. In contrast, missing data in $F$-statistics is most commonly handled by estimating a standard error by resampling along the genome \citep{patterson2012}, and so missing data results in larger standard error.

@@

\subsubsection{Individual vs. population-based analyses}
The final issue is that PCA is commonly run on individual-based data, whereas $F$-statistics often group individuals into populations. However, population-based PCA has been the default in the past \citep{cavalli-sforza1994}, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}. Often, an individual-based PCA is used to justify grouping individuals into populations; i.e. individuals that form a tight cluster on a PCA-plot have similar relationships to everyone else in the data set, and can thus be treated as a unit of analysis. Thus, if the assumptions are satisfied, $F$-statistics for individual-based and population-based analyses are expected to be very similar. PCA, on the other hand, is strongly impacted by the number of individuals from each population \citep[e.g.][]{mcvean2009}; as each individual is weighted equally, variation related to populations with many samples will be overrepresented on the first PCs.

\subsubsection{Summary}
Motivated by $F$-statistics, the PCAs I consider here are based on estimated population allele frequencies, whereas most genome-scale studies of human genetic variation use observed individual-based allele frequencies that are normalized by overall allele frequencies. Thus, some care will be required to directly extend the interpretations developed here to individual-based PCAs. However, the differences are largely due to conventions, and particularly for studies where the description of population structure is a major focus, results might be easier to interpret if conventions regarding missing data, normalization and estimation of allele frequencies are used consistently between $F$-statistics and PCA.

\subsection{The Apportionment of Human Diversity}
Most genetic variation in humans is shared between all of us, but the around 15\% that can be explained by population structure can be leveraged to study our history and diversity in great detail \citep{dobzhansky1972, cavalli-sforza1994, novembre2022}. For some data sets it is possible to predict an individuals' origin at a resolution of a few hundred kilometers \citep{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.

@@

From a practical perspective, formulating hypotheses and designing studies in terms of discrete populations with ``uniform'' genetic backgrounds is often sensible, as it enables  e.g. prediction of phenotypes \citep{berg2019, yair2021}, inference of demographic parameters, and schematic models of human genetic history \citep{patterson2012}. In a similar vein, when interpreting $F$-statistics in the context of admixture graphs, we make the  implicit assumtion that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
However, these simplifications do come at a cost, both in terms of model violations that may invalidate statistical results, and in terms of deemphasizing people that do not rigidly fall into predefined genetic groups.
